### 2023-04-28 chapter2/chapter2.tex
Updated discussion outline
Added sentence about analogy to Pfam and references
Removed redundant use of the word feature

However, no known subsequent studies have determined if similar patterns of conservation are found in the IDRs of other systems. As another foundational model organism with abundant genomic information across many evolutionary lineages~\cite{Yang2018, Miller2018, Kim2021, Gramates2022}, the fruit fly, \textit{Drosophila melanogaster}, is a natural choice for subsequent investigation. Furthermore, given its complex multicellular development process and shared signaling pathways with humans, the findings of such a study would significantly advance our understanding of the role of IDRs in gene regulation as well as human health and disease. The concordance of these results with the previously identified IDR clusters would also have profound implications for the broader mechanisms of IDR evolution. For example, the absence of global patterns of evolutionary signatures across IDRs in \textit{Drosophila} would suggest they are property of IDRs which is unique to yeast. In contrast, the identification of clusters similar to those in yeast would indicate the existence of a taxonomy of IDRs which is conserved across the tree of life.
% The latter result would represent a significant step towards the creation of resources for the classification of IDRs analogous to those for folded domains such as Pfam, CATH or SCOP.

Therefore, in this work we apply a series of phylogenetic models to a set of orthologous IDRs identified in the \textit{Drosophila} genome to dissect the forces driving their evolution. Our analyses span multiple levels, ranging from the sequences that compose these regions to the distributed features that characterize them as a whole. For the latter, though the previous approach relied on simulations to generate the null distribution for a hypothesis of no constraint, we instead leverage a fully statistical phylogenetic comparative framework~\cite{Cornwell2017}. By comparing models of constrained and unconstrained continuous trait evolution, \textit{i.e.} the Brownian motion and Ornstein-Uhlenbeck models, respectively, we can demonstrate evidence of selective constraint on features independent of any assumptions about the underlying process of sequence evolution. However, we also explore hybrid approaches that combine simulations with phylogenetic comparative methods to test increasingly refined models of IDR evolution. We find that IDRs exhibit unique patterns of amino acid substitution and that in some proteins disorder itself is a dynamically evolving property. Furthermore, though IDRs are broadly unconstrained along several axes of feature evolution, we identify specific clusters of IDRs with shared patterns of feature constraint. As in yeast, these clusters are enriched for proteins with specific functional annotations, which suggests the preservation of distributed features and their associated functions is a widespread mechanism of IDR evolution.

@@

% II. Exchangeability and rate matrices
%    A. Reproduced trends seen in previous substitution models fit to disorder regions
%    B. Interpretations of exchangeability and rate matrices
%        1. Disorder-enriched residues are less interchangeable and play specific roles in these regions
%        2. Disorder-depeleted residues not constrained by packing and can act more generically as small hydrophobic patches
%            i. Stickers and spacers from Holehouse
%            ii. Staller papers and other high throughput mutational scanning of ADs
%            iii. Halfmann glutamine vs asparagine paper
%        3. Note about variances of parameters as shown in CV graphs

% III. Disorder score analyses
%    A. The analysis of features and disorder scores is effectively a natural ``mutational scanning'' experiment [citation to Eddy?]
%        1. Contrasts of features and disorder scores shows the disorder predictor is sensitive to most features used in this analyses
%        2. However, highest correlations to expected features like fraction_disorder and hydropathy
%        3. Interestingly, negatively correlated with wf_complexity. Many disordered regions have low levels of sequence complexity. However this is a statistical rather than a biophysical criterion, and some low complexity regions are structured, like collagen. This indicates that these models can confuse correlation and causation during training, and their representations should be interpreted carefully.
%        4. Furthermore, resolution of analyses limited because features calculated at level of regions, whose lengths can exceed 1,000 residues, rather than a fixed window around each residue.
%        5. Disorder scores are weakly correlated with many motifs. Furthermore, unclear if predictor is responding directly to these motifs as a signal of disorder, or they appear in regions which evolving towards disorder. Interestingly docking (DOC) and ligand binding (LIG) sites tend to have negative correlations whereas the modification (MOD) sites have positive correlations. Docking and ligand binding sites are generally mediated by small hydrophobic patches, so the predictor could be responding to increased hydrophobicity caused by additional of hydrophobic residue ``completing'' the motif rather than the motif itself. Likewise, phosphorylation sites are highly enriched in IDRs and many are targeted to disorder-promoting residues like serine or threonine. Thus, it's unclear if the predictor is responding directly to these features or features which are correlated with them. More in depths of the representations of these disorder predictors would shed led on these questions.
%    B. GO annotation enrichment analysis suggests disorder scores undergo rapid evolution in proteins involved in DNA repair and found in extracellular structure
%        1. Not corrected for multiple testing nor is the false discovery rate controlled, so should be considered as hypotheses for further investigation
%        2. However, this result is at least consistent with other known functions of disorder
%            i. Interactions with extracellular components and signaling, both highly enriched in disorder?
%            ii. These regions could contain MoRFs, so they are already on the boundary between folded and disorder
%            iii. As evolution optimizes these interactions, small changes in the biophysical properties of these regions may have disproportionate effects on their structural ensembles
%            iv. The variable scoring by the disorder predictor may reflect this kind of ``phrase transition''

% IV. Feature analyses
%    A. Though no discrete clusters are visible, only captures x% of the overall variation, so significant amounts of variation not represented in these axes
%    B. Hierarchical clustering of the rates reveals additional structure.
%    C. Other non-linear dimensionality reductions like UMAP be more successful at preserving the fine details of the high-dimensional distributions in the low-dimensional space
%        1. However, these methods designed primarily for visualization and as result can significant distort the global structure of high-dimensional data to preverse its local topology. Thus, these embeddings are generally not suitable for downstream analyses.

% V. OU analyses
%    A. Discussion of issues with statistical inference and need for simulations to address inference issues with OU model
%        1. Lack of power or prone to type I errors?
%    B. Better models for handling motif counts
%        1. Unlike other features where small changes in a sequence general yield small changes in their value, the complexity of the regular expressions made them highly sensitive to small perturbations.
%    C. Need for comparisons with simulations with amino acid substitution and indel models
%        1. Weaknesses are explicit assumptions about substitution and indel processes and high computational cost
%        2. Is also a strength because allows testing of increasingly specific hypotheses for IDR evolution
%        3. Complementary approach -- fewer assumptions about description of evolution but cannot be tailored to specific hypotheses
%        4. Hybrid approach using comparisons of rates as the statistical test and simulations to generate the null distribution
%    D. Mention other distribution free approaches

>>>>

However, no known subsequent studies have determined if similar patterns of conservation are found in the IDRs of other systems. As another foundational model organism with abundant genomic information across many evolutionary lineages~\cite{Yang2018, Miller2018, Kim2021, Gramates2022}, the fruit fly, \textit{Drosophila melanogaster}, is a natural choice for subsequent investigation. Furthermore, given its complex multicellular development process and shared signaling pathways with humans, the findings of such a study would significantly advance our understanding of the role of IDRs in gene regulation as well as human health and disease. The concordance of these results with the previously identified IDR clusters would also have profound implications for the broader mechanisms of IDR evolution. For example, the absence of global patterns of evolutionary signatures across IDRs in \textit{Drosophila} would suggest they are property of IDRs which is unique to yeast. In contrast, the identification of clusters similar to those in yeast would indicate the existence of a taxonomy of IDRs which is conserved across the tree of life. The latter result would represent a significant step towards the creation of resources for the classification of IDRs analogous to those for folded domains such as Pfam~\cite{Mistry2020}, CATH~\cite{Sillitoe2020}, or SCOP~\cite{Andreeva2013, Andreeva2019}.

Therefore, in this work we apply a series of phylogenetic models to a set of orthologous IDRs identified in the \textit{Drosophila} genome to dissect the forces driving their evolution. Our analyses span multiple levels, ranging from the sequences that compose these regions to the distributed features that characterize them as a whole. For the latter, though the previous approach relied on simulations to generate the null distribution for a hypothesis of no constraint, we instead leverage a fully statistical phylogenetic comparative framework~\cite{Cornwell2017}. By comparing models of constrained and unconstrained continuous trait evolution, \textit{i.e.} the Brownian motion and Ornstein-Uhlenbeck models, respectively, we can demonstrate evidence of selective constraint on features independent of any assumptions about the underlying process of sequence evolution. However, we also explore hybrid approaches that combine simulations with phylogenetic comparative methods to test increasingly refined models of IDR evolution. We find that IDRs exhibit unique patterns of amino acid substitution and that in some proteins disorder itself is a dynamically evolving property. Furthermore, though IDRs are broadly unconstrained along several axes of feature evolution, we identify specific clusters of IDRs with shared patterns of constraint. As in yeast, these clusters are enriched for proteins with specific functional annotations, which suggests the preservation of distributed features and their associated functions is a widespread mechanism of IDR evolution.

@@

% II. Exchangeability and rate matrices
%    A. Interestingly, the exchangeability matrices suggest that although there is net flux towards disorder by depleting structure-promoting residues, disordered residues are more biochemically distinct in disordered regions than in structured regions. This suggests that although disorder-promoting amino acids are often grouped into a single category and described as largely hydrophilic, they have distinct roles in disordered regions. There is evidence that highly similar residues such as glutamine and asparagine have distinct properties in disordered regions that yielded different biophysical characters of phase-separated condensates, so this aligns with other observations. Conversely, structure promoting residues are more interchangeable. One hypothesis is as these bulkier hydrophobic residues are not constrained by packing, they are free to act more generically as small hydrophobic patches that mediate interactions in SLiMs and other binding regions in IDRs. Another model for this behavior is the stickers and spacers model of phase separation.
%    A. A difficulty with fitting substitution models is their dependence on sequence alignments which were created using scoring matrices that are in turn often derived from substitution models. Some approaches attempt to minimize the impact of this circular dependency through an EM-like procedure where substitution models are fit and the sequences in alternating rounds until convergence. Because our analyses fit the substitution models directly from alignments, the observed patterns we observe could be artifacts of using a scoring matrix largely fit to folded domains to align IDRs
%    A. However, as the exchangeability and rate matrices broadly reproduced the trends observed in previous studies which fit amino acid substitution models to disorder regions, these observations are likely not artifactual but instead represent reproducible differences in the substitution rates between residues.
%    B. Interpretations of exchangeability and rate matrices
%        1. Disorder-enriched residues are less interchangeable and play specific roles in these regions
%        2. Disorder-depeleted residues not constrained by packing and can act more generically as small hydrophobic patches
%            i. Stickers and spacers from Holehouse
%            ii. Staller papers and other high throughput mutational scanning of ADs
%            iii. Halfmann glutamine vs asparagine paper
%        3. Note about variances of parameters as shown in CV graphs

% III. Disorder score analyses
%    A. The analysis of feature and score contrasts is effectively a natural ``mutational scanning'' experiment. Though the mutations generated by evolution are not a random or exhaustive sample of sequence space, they are perturbations around a common ancestral sequence which can reveal the relationship between a region's biophysical properties and its propensity for disorder. Accordingly, we find disorder scores have the strongest correlations with features that measure a region's overall hydrophobicity, \textit{e.g.} fraction_disorder and hydropathy. As these features are known correlates of intrinsic disorder, this relationship is expected. However, the strength of these correlations indicate that a region's relative proportion of hydrophilic and hydrophobic residues is, to a first approximation, the largest determinant of its structural state.
%        3. Excluding these hydrophobicity-related features, the second strongest association is a negatively correlation with wf_complexity, indicating the predictor strongly associates low complexity with intrinsic disorder. However, sequence complexity is a statistical rather than a biophysical criterion, and while many disordered regions have low levels of complexity, some low-complexity regions, like collagen, are structured. This suggests that while disorder predictors are in general accurate classifiers of a residue's structural state, they can conflate correlation and causation, and therefore in some cases their predictions may require careful interpretation.
%
%        5. Interestingly, the disorder scores have significant, if weak, correlations with most features used in this analysis, including the motifs. Disorder scores are weakly correlated with many motifs. Furthermore, unclear if predictor is responding directly to these motifs as a signal of disorder, or they appear in regions which evolving towards disorder. Interestingly docking (DOC) and ligand binding (LIG) sites tend to have negative correlations whereas the modification (MOD) sites have positive correlations. Docking and ligand binding sites are generally mediated by small hydrophobic patches, so the predictor could be responding to increased hydrophobicity caused by additional of hydrophobic residue ``completing'' the motif rather than the motif itself. Likewise, phosphorylation sites are highly enriched in IDRs and many are targeted to disorder-promoting residues like serine or threonine. Thus, it's unclear if the predictor is responding directly to these features or features which are correlated with them. More in depth analyses of the representations of these disorder predictors would shed led on these questions.
%        1. In, the disorder scores have a significant relationship with most features used this analysis.
%        4. Furthermore, resolution of analyses limited because features calculated at level of regions, whose lengths can exceed 1,000 residues, rather than a fixed window around each residue.

%    B. The GO annotation enrichment analysis indicate that proteins involved in DNA repair and the formation of extracellular structure contain disproportionate number of regions whose disorder scores are rapidly evolving. Because the significance tests were not corrected for multiple testing or controlled for false discovery rate, we caution against over-interpreting this result and instead consider as a hypothesis for further investigation. In general, however, these regions with rapidly evolving disorder scores may correspond to molecular recognition features (MoRFs), which are modules in IDRs that undergo a disorder-to-order transition on binding their targets. Because MoRFs already exist on the boundary between disorder and structure, small changes in the biophysical properties of these regions may have disproportionate effects on their structural ensembles. Thus, the most variable disorder scores may reflect instances where a mutation triggered a ``phase transition'' between largely structured or disordered native state.

% IV. Feature analyses
%    A. In the PCAs of the feature roots and rates, though no discrete clusters are visible in the three largest components, they only accounted for X\% of the total variation between region. This suggests, a majority of variation is contained other in these components.
%    B. The success of the hierarchical clustering of the rates at revealing additional structure suggests
%    C. Other non-linear dimensionality reductions like UMAP be more successful at preserving the fine structure of these high-dimensional distributions in the low-dimensional space. However, as these methods are designed primarily for visualization, they can significant distort the global structure of high-dimensional data. Thus, these embeddings are generally not suitable for downstream analyses. Instead, semi-supervised approaches for feature learning may yield more meaningful representations. In fact, a recent study used a semi-supervised learning approach that identified orthologous IDRs to learn novel disorder-associated features. Though they used a simple neural network that aided in interpretability, other similar semi-supervised approaches have used more complex networks to learn rich representations of protein regions. Similar approaches may yield representations of the space of disordered proteins that are more clearly visualized and manipulated.

% V. OU analyses
%    A. While preliminary analyses indicate
%    A. Discussion of issues with statistical inference and need for simulations to address inference issues with OU model
%        1. Lack of power or prone to type I errors?
%    B. Better models for handling motif counts
%        1. Unlike other features where small changes in a sequence general yield small changes in their value, the complexity of the regular expressions made them highly sensitive to small perturbations.
%    C. Need for comparisons with simulations with amino acid substitution and indel models
%        1. Weaknesses are explicit assumptions about substitution and indel processes and high computational cost
%        2. Is also a strength because allows testing of increasingly specific hypotheses for IDR evolution
%        3. Complementary approach -- fewer assumptions about description of evolution but cannot be tailored to specific hypotheses
%        4. Hybrid approach using comparisons of rates as the statistical test and simulations to generate the null distribution
%    D. Mention other distribution free approaches
